---
title: "Introduction to R Homework Answers"
# Unlinked from the sidebar and excluded from site search so students do not
# stumble onto the answers. The page is still published, so the URL can be shared
# deliberately once the homework has been discussed.
search: false
author:
  - "Gabriel Mateus Bernardo Harrington"
  - "Matthew Bracher-Smith"
date: "2026-10-01"
format:
  html:
    number-sections: true
editor: source
execute:
  echo: true
---

```{r}
#| include: false
library(tidyverse)
```

## R basics

### 1. Calculations

```{r}
sqrt(6 * 2)
4 + 3 - 2
1046 * 934
```

### 2. The `women` dataset

`women` ships with base R, in the `datasets` package, so no package needs loading.

```{r}
data(women)
nrow(women)
colnames(women)
```

```{r}
women$ages <- sample(18:90, nrow(women), replace = TRUE)
head(women)
```

```{r}
colMeans(women)
colSums(women)

# or one function at a time over every column
sapply(women, mean)
```

`summary()` gives the minimum, quartiles, median, mean and maximum of every column
at once, which is a much better first look at a dataset than a single mean.

```{r}
summary(women)
```

### 3. Reading, subsetting and writing

```{r}
#| message: false
cohort_base <- read.table(
  "http://tbb.bio.uu.nl/BDA/fig4.tsv",
  sep = "\t",
  header = TRUE
)
cohort <- read_tsv("http://tbb.bio.uu.nl/BDA/fig4.tsv")

cohort_base
cohort
```

`read_tsv()` returns a tibble, which prints each column's type and only as many
rows as fit, and it needs no `sep` or `header` arguments. `read.table()` needs
telling both, and returns a plain data frame.

```{r}
females <- cohort[cohort$Gender == "F", ]
# or, with dplyr
females <- filter(cohort, Gender == "F")
females
```

```{r}
#| eval: false
write_csv(females, "fig4-females.csv")
```

## Data types and structures

### 4. Types, vectors, matrices and data frames

R's basic types are numeric (double), integer, character, logical, complex and
raw. They are held in vectors, matrices, arrays, lists, factors and data frames.

```{r}
a <- c(1, 2, 5.3, 6, -2, 4) # numeric vector
b <- c("one", "two", "three") # character vector
c <- c(TRUE, TRUE, TRUE, FALSE, TRUE, FALSE) # logical vector
a
b
c
c[c(1, 3)]
```

```{r}
mymatrix <- matrix(
  1:16,
  nrow = 4,
  ncol = 4,
  byrow = TRUE,
  dimnames = list(c("R1", "R2", "R3", "R4"), c("C1", "C2", "C3", "C4"))
)
mymatrix
```

`dimnames` takes a list of length two: the row names first, then the column names.
Either can be `NULL` to leave that dimension unnamed.

Note that the data has to fit the matrix: `matrix(1:20, nrow = 4, ncol = 4)` warns
that the data length differs from the size of the matrix, because 20 values do not
fit in 16 cells.

```{r}
mydata <- data.frame(
  id = 1:4,
  colour = c("red", "white", "red", NA),
  passed = c(TRUE, TRUE, TRUE, FALSE)
)
mydata

names(mydata) <- c("ID", "Colour", "Passed")
mydata
```

### 5. Packages

Any package will do. For example:

```{r}
#| eval: false
install.packages("skimr")
library(skimr)
skim(women) # a readable summary of every column, with small histograms
```

## Tibbles and readr

### 6. Making a tibble

```{r}
tib <- tibble(
  name = c(
    "Luke", "Julia", "Matt", "Simon", "Laura", "Paul", "Frank",
    "Chris", "Pam", "Sue", "Sim", "Tom", "Ben"
  ),
  age = c(22, 21, 18, 19, 20, 16, 17, 26, 23, 25, 28, 18, 19),
  height = c(182, 180, 178, 183, 180, 179, 176, 175, 180, 181, 182, 170, 175),
  gender = c("M", "F", "M", "M", "F", "M", "M", "M", "F", "F", "M", "M", "M")
)
tib
```

```{r}
# 6a
tib[4, ]

# 6b
tib$height[tib$name == "Luke"]

# 6c
sum(tib$gender == "M") / sum(tib$gender == "F")

# 6d
mean(tib$age)
sd(tib$age)
```

### 7. `iris` as a tibble

```{r}
as_tibble(iris)
```

### 8. Why tibbles?

They never change the type of your input, never create row names, only recycle
vectors of length one, allow list-columns and non-standard column names, and print
only what fits on screen along with each column's type. Each of those removes a way
for a data frame to surprise you.

### 9. Useful readr options

`col_names`, `col_types`, `skip`, `n_max`, `na` (which values count as missing),
`comment`, and `locale` (decimal mark and encoding).

### 10 and 11. The unclean phenotype file

```{r}
pheno <- read_tsv(
  "https://raw.githubusercontent.com/H-Mateus/Data-for-Life-Sciences-in-R/main/data/pheno_unclean.txt",
  show_col_types = FALSE
)
pheno
summary(pheno)
```

The phenotype column is meant to be coded 0/1, but its maximum is 5. The column
names contain spaces and punctuation, so they need backticks:

```{r}
filter(pheno, `Phenotype (coded 0/1)` > 1)
```
