---
title: "Data Wrangling in R II Homework Answers"
# Unlinked from the sidebar and excluded from site search so students do not
# stumble onto the answers. The page is still published, so the URL can be shared
# deliberately once the homework has been discussed.
search: false
author:
  - "Matthew Bracher-Smith"
  - "Gabriel Mateus Bernardo Harrington"
date: "2026-10-12"
format:
  html:
    number-sections: true
    highlight: tango
editor: source
execute:
  echo: true
---

```{r}
#| include: false
library(tidyverse)
library(conflicted)
conflicts_prefer(dplyr::filter)
```

```{r}
#| message: false
base_path <- here::here("Simulated_data", "generated_uk")

baseline  <- read_csv(file.path(base_path, "bioinformatics_msc_stats_baseline.csv"),
                      show_col_types = FALSE)
visits    <- read_csv(file.path(base_path, "bioinformatics_msc_stats_visits.csv"),
                      show_col_types = FALSE)
mortality <- read_csv(file.path(base_path, "bioinformatics_msc_stats_mortality_comorbidity.csv"),
                      show_col_types = FALSE)
```

## 1. The join commands, and the two protective arguments

The mutating joins all take the form `*_join(x, y, by = ...)`:

- `left_join()` — keeps every row of `x`, fills with `NA` where `y` has no match. The default choice.
- `inner_join()` — keeps only rows present in both.
- `right_join()` — keeps every row of `y`.
- `full_join()` — keeps every row of both.

The filtering joins do not add columns, they only subset `x`:

- `semi_join()` — keeps rows of `x` that *have* a match in `y`.
- `anti_join()` — keeps rows of `x` with *no* match. Excellent for finding what is missing.

The two arguments that prevent mistakes:

- **`by`** — always state the key explicitly. A natural join on every shared column name is rarely what you want, and it changes silently when someone adds a column.
- **`relationship`** — declare what you expect (`"one-to-one"`, `"many-to-one"`, and so on). dplyr then errors if the data disagrees, instead of quietly producing extra rows.

## 2. Joining baseline to mortality

```{r}
joined <- left_join(baseline, mortality, by = "participant_id")

c(baseline = ncol(baseline), mortality = ncol(mortality), joined = ncol(joined))
```

### 2a.

`r ncol(baseline)` + `r ncol(mortality)` would be `r ncol(baseline) + ncol(mortality)`, but the result has `r ncol(joined)`. The key `participant_id` appears once rather than twice, so we lose one column.

### 2b.

```{r}
sum(grepl("\\.x$|\\.y$", names(joined)))
```

The two tables share `r length(intersect(names(baseline), names(mortality))) - 1` column names besides the key — `age_years`, `sex_at_birth`, `frailty_category` and so on. dplyr cannot have two columns with the same name, so it suffixes both copies.

What to do about it: `select()` only the columns you actually need from the second table before joining. Carrying 20 duplicated columns around is how you end up analysing `age_years.y` by accident.

```{r}
tidy_join <- baseline |>
  left_join(select(mortality, participant_id, died_during_followup, age_at_death),
            by = "participant_id")

ncol(tidy_join)
```

## 3. The join that changes your row count

### 3a and 3b.

```{r}
visit2 <- filter(visits, visit_number == 2)

joined_v2 <- left_join(baseline, visit2, by = "participant_id")

c(before = nrow(baseline), after = nrow(joined_v2))
```

You would expect 2,400 — one row per participant. You get **2,402**.

### 3c.

```{r}
visit2 |>
  count(participant_id) |>
  filter(n > 1)
```

Two participants have their visit-2 record entered **twice**. They are exact duplicates:

```{r}
dupes <- visits |>
  filter(participant_id %in% c("BIO00010", "BIO00018"), visit_number == 2)

c(rows = nrow(dupes), distinct_rows = nrow(distinct(dupes)))
```

Each of those two participants therefore matches two rows in `visit2`, and `left_join()` duly returns both.

### 3d.

```{r}
#| error: true
left_join(baseline, visit2, by = "participant_id", relationship = "one-to-one")
```

It errors immediately and names the offending row. This is a better failure because it happens **at the join**, where the cause is obvious, rather than surfacing three steps later as a mean that is slightly wrong in a way nobody can explain. Two extra rows in 2,400 will not look wrong in any summary you produce.

## 4. Reshaping the weights

### 4a.

```{r}
#| warning: true
wide_bad <- visits |>
  select(participant_id, visit_number, weight_kg) |>
  pivot_wider(names_from = visit_number, values_from = weight_kg, names_prefix = "visit_")

class(wide_bad$visit_2)
```

The columns are **list-columns**, not numeric, and `pivot_wider()` warns that values are "not uniquely identified".

### 4b.

The same two duplicated rows from question 3. `pivot_wider()` expects one value per participant per visit; for `BIO00010` and `BIO00018` at visit 2 it finds two, and rather than guess it wraps every cell in a list.

### 4c.

Because it is **loud**. `pivot_wider()` warns and hands back an obviously wrong data type, so you find out within seconds. `left_join()` said nothing at all. The same data problem, discovered immediately in one case and potentially never in the other — which is the whole argument for checking row counts yourself rather than relying on the function to tell you.

### 4d.

Since the duplicates are exact, `distinct()` is a defensible fix — we are removing a double-entered record, not choosing between two conflicting measurements:

```{r}
wide <- visits |>
  select(participant_id, visit_number, weight_kg) |>
  distinct() |>
  pivot_wider(names_from = visit_number, values_from = weight_kg, names_prefix = "visit_")

c(rows = nrow(wide), cols = ncol(wide))
class(wide$visit_2)
```

Had the two rows *disagreed* on weight, `distinct()` would not have helped and the right answer would have been to go back to the source and ask which is correct. Record whichever you did, and why.

## 5. Back to long format

```{r}
long <- wide |>
  pivot_longer(
    -participant_id,
    names_to = "visit",
    values_to = "weight_kg",
    values_drop_na = TRUE
  )

c(rows = nrow(long), if_complete = 2400 * 6)
```

`r format(nrow(long), big.mark = ",")` rather than 14,400, because the study has attrition: participants miss visits, withdraw, or die, so most people do not have all six measurements. `values_drop_na = TRUE` discards those empty cells.

## 6. `across()`

### 6a.

```{r}
baseline |>
  summarise(across(c(bmi_kg_m2, systolic_bp_mmHg), ~ mean(.x, na.rm = TRUE)))
```

### 6b.

```{r}
baseline |>
  summarise(
    across(
      c(bmi_kg_m2, systolic_bp_mmHg),
      list(mean = ~ mean(.x, na.rm = TRUE), median = ~ median(.x, na.rm = TRUE))
    )
  )
```

### 6c.

```{r}
baseline |>
  group_by(imd_quintile) |>
  summarise(
    across(c(bmi_kg_m2, systolic_bp_mmHg), ~ round(mean(.x, na.rm = TRUE), 1)),
    n = n()
  )
```

There are **six** groups, not five. IMD quintiles run 1 to 5, so the single participant coded `9` is an invalid value — most likely a data entry error or a missing-value code that was never translated.

Grouping is a good way to find these: an unexpected group with a tiny `n` almost always means a coding problem rather than a real subgroup. Note also that a group of one gives you a "mean" with no spread at all, which will look perfectly reasonable in a summary table.
