---
title: "Data Wrangling in R II Homework"
author:
  - "Matthew Bracher-Smith"
  - "Gabriel Mateus Bernardo Harrington"
date: "2026-10-12"
format:
  html:
    number-sections: true
    highlight: tango
editor: source
execute:
  echo: true
---

This document contains all questions for **Data Wrangling in R II**. Please create a Quarto document containing all text, code and output used to answer the questions.

## The data

We will use the module cohort — a simulated study of 2,400 older adults, in three linked tables. You can read them straight from the web:

```{r}
#| eval: false
library(tidyverse)

base_url <- "https://raw.githubusercontent.com/H-Mateus/Data-for-Life-Sciences-in-R/main/Simulated_data/generated_uk/"

baseline  <- read_csv(paste0(base_url, "bioinformatics_msc_stats_baseline.csv"))
visits    <- read_csv(paste0(base_url, "bioinformatics_msc_stats_visits.csv"))
mortality <- read_csv(paste0(base_url, "bioinformatics_msc_stats_mortality_comorbidity.csv"))
```

All three share a `participant_id` column. `visits` has one row per participant per visit, at 0, 12, 24, 36, 48 and 60 months.

## Joins

1. List and describe the tidyverse commands used to join two datasets together. What two key arguments can you use to prevent mistakes in your joins?

2. Join `baseline` to `mortality` by `participant_id`.

    * 2a. How many columns did each table have, and how many does the result have? It is *not* the sum — explain the difference.
    * 2b. How many columns in the result end in `.x` or `.y`? What causes that, and what would you do about it?

3. **Predict before you run.** Join `baseline` to only the visit-2 rows of `visits`, by `participant_id`:

    * 3a. How many rows does `baseline` have? How many do you *expect* after the join?
    * 3b. Run it. How many rows do you actually get?
    * 3c. Find the participants responsible. What is wrong with those rows?
    * 3d. Re-run the join with `relationship = "one-to-one"`. What happens? Why is this a better failure than the one in 3b?

::: {.callout-important}
Question 3 is the single most important thing in this homework. A join that silently changes your row count will quietly bias every mean, count and model you calculate afterwards, and nothing will warn you. Get into the habit of checking `nrow()` before and after **every** join.
:::

## Reshaping

4. Reshape the weight measurements so there is one row per participant and one column per visit:

```{r}
#| eval: false
visits |>
  select(participant_id, visit_number, weight_kg) |>
  pivot_wider(names_from = visit_number, values_from = weight_kg, names_prefix = "visit_")
```

  * 4a. Run it. R gives you a warning and the columns are not the type you would expect. What type are they, and what is the warning telling you?
  * 4b. What is the underlying cause? Compare your answer with question 3 — **the same two rows are responsible for both**.
  * 4c. Why is this failure easier to catch than the one in question 3?
  * 4d. Fix the input so that `pivot_wider()` produces ordinary numeric columns, and say what you decided to do about the offending rows and why.

5. Take your fixed wide table from 4d and use `pivot_longer()` to put it back into long format, dropping missing values. How many rows do you get, and why is it fewer than 2,400 × 6?

## Applying a function across columns

6. Using `baseline`:

    * 6a. Use `across()` inside `summarise()` to get the mean `bmi_kg_m2` and `systolic_bp_mmHg` in one command, ignoring missing values.
    * 6b. Extend it to report both the mean and the median for each.
    * 6c. Group by `imd_quintile` first, and add a count of participants per group. Does anything about the groups look wrong to you?

**Suggested Reading**

- [R for Data Science 2e](https://r4ds.hadley.nz/), chapters 5 (data tidying) and 19 (joins)
- The docs for [tidyr](https://tidyr.tidyverse.org/) and [dplyr](https://dplyr.tidyverse.org/)
- [`vignette("pivot", package = "tidyr")`](https://tidyr.tidyverse.org/articles/pivot.html)
- Strings and regular expressions are no longer taught in a session, but there is a [reference page](../10_resources/02_reference/strings_and_regex.qmd) with worked examples and practice if you need them.
