---
title: "Data Wrangling in R I Homework Answers"
# Unlinked from the sidebar and excluded from site search so students do not
# stumble onto the answers. The page is still published, so the URL can be shared
# deliberately once the homework has been discussed.
search: false
author: "Matthew Bracher-Smith"
date: "2024-10-17"
format: 
  html:
    number-sections: true
    highlight: tango
editor: source
execute:
  echo: true
  output: false
---

This document contains all questions for the session 'Data Wrangling in R I'. Please create a Quarto document containing all text, code and output used to answer the questions.

```{r}
#| include: false
library(readr)
library(dplyr)
library(stringr)
data(starwars)
```

## Quarto

1. Hand this homework in as a Quarto document, rendered to html. Before the dplyr questions, write a paragraph explaining the tidyverse concept. Your paragraph should include:

    - Example code
    - Inline code
    - A blockquote
    - A list
    - Bold or italic text

Check the rendered html rather than the source: inline code should show its *result*, and the blockquote, list and emphasis should all render as formatting rather than as literal symbols.

## Tidy Data
2. What 3 main rules do we need to follow for data to be in tidy format?

The 3 main rules for tidy data are:

    - Each observation must have its own row
    - Each variable must have its own column
    - Each value must have its own cell

3. Load `readr` and use it to read in the dataset at "http://stat405.had.co.nz/data/pew.txt". You should have a tibble with 18 rows and 11 columns showing data on the relationship between religion and income in the US. Is the data in tidy format? Explain why.

No, it's not in tidy format.

    - This is because it has values as column names. For example, "<$10k" is a value but has
      been listed as a column header.
    - A tidy dataset would instead have three columns called religion, income and frequency

4. Look at [the paper by Hadley Wickham](https://www.jstatsoft.org/article/view/v059i10/) describing tidy data. Section 3 outlines how to turn messy datasets into tidy ones. Briefly state the 5 most common problems that make a dataset messy and the solutions Hadley proposes.

The 5 most common problems, followed by their solutions are:

    - Column headers are values, not variable names - solution is to *melt* the dataframe
    - Multiple variables are stored in one column - solution is to *separate* data into new columns
    - Variables are stored in both rows and columns - solutions is to *cast* data after melting
    - Multiple types of observational units are stored in the same table - solution is to move
      observational units to separate datasets
    - A single observational unit is stored in multiple tables - add columns for file names
      and *merge* datasets

## dplyr
These exercises require use of the dplyr verbs we have learned so far. Some questions will require small variations on these that you need to look up; you may find it especially useful to check the documentation on [**scoped variants of the standard verbs**](https://dplyr.tidyverse.org/reference/scoped.html), or the recent equivalents in [**pick**](https://dplyr.tidyverse.org/reference/pick.html) and [**across**](https://dplyr.tidyverse.org/reference/across.html) (we will review both options in the next session). All tasks that require use of more than one verb should be done using the pipe. Show the output from each question in a new cell, where a single paragraph of pipes is used to answer each question.

If you're struggling with a question that requires a lot of steps, try to sketch out the bones of the code before filling in the details. For instance, if you're asked to show the European countries in 1990, sorted by life expectancy, you might first write out the basic order of things, like so:
```{r}
#| eval: false
gapminder |>
  filter() |>
  select() |>
  arrange()
```

**read in the dataset at 'http://stat405.had.co.nz/data/weather.txt' using `readr`**

```{r}
#| warning: false
weather <- read_tsv('http://stat405.had.co.nz/data/weather.txt')
```

5. convert all column names to title case, except 'id', which should be all capitals

```{r}
#| warning: false
weather |>
  rename_all(str_to_title) |>
  rename(ID = Id)

# could also have selected everything but 'id' for the first rename
# this isn't necessary though
weather |>
  rename_at(vars(-id), funs(str_to_title)) |>
  rename(ID = id)
```

```{r}
# as above, but using rename_with
weather |>
  rename_with(str_to_title, everything()) |>
  rename(ID = Id)
```

6. choose columns ID, Year, Month and d1 to d10. Use `num_range` to select the columns d1:d10

```{r}
weather |>
  select(id, year, month, num_range('d', 1:10))
```

**read in the dataset at 'https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv' using `read_csv()`. This data contains weights and sleep times for mammals. You should have 11 columns and 83 rows**

```{r}
#| message: false
sleep <- read_csv('https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv')
```

7. select the name and genus columns, and all columns ending with 'wt'. Remove all rows with missing values,then print the first 20 rows from the final dataframe.

```{r}
sleep |>
  select(name, genus, ends_with('wt')) |>
  na.omit() |>
  print(n = 20)
```

8. show the columns name, order, sleep_total and awake for all animals in the order 'Artiodactyla', sorted by descending sleep time.

```{r}
sleep |>
  filter(order == 'Artiodactyla') |>
  select(name, order, sleep_total, awake) |>
  arrange(desc(sleep_total))
```

9. doubles should never be compared using `==`. Instead, use `dplyr::near()` to keep rows with 'sleep_total' equal to 9.4 and select columns containing the string 'or' anywhere in their names

```{r}
# make sure they have used near() and not '=='
# make sure used contains() and not starts_with() or ends_with()
sleep |>
  filter(dplyr::near(sleep_total, 9.4)) |>
  select(contains('or'))
```

**Load the starwars dataset**

```{r}
#| warning: false
data("starwars")
```

10. filter hair_color to keep those rows *containing* brown (including combinations like 'brown, grey') **or** eye_color that is brown *only*, then select the column range from 'name' to 'eye_color', and the columns 'gender', 'homeworld' and 'species.' Next, create a new boolean column called 'male_brunette', which is TRUE only for males with exclusively brown hair. Sort by descending height and re-order the columns, using `select()` and `everything()`, to put 'male_brunette' directly after the 'name' column. Finally, replace underscores in the column names with spaces, change all instances of 'color' to 'colour', and make all column names title case using `str_to_title()`. Print the top 5 rows only.

```{r}
# make sure the filter uses grepl (or str_detect) for hair colour, but '==' for eye colour
# need to have used an or (|) for the filter, not an and (& or ,)
# select could also have been done by excluding the missing columns with '-'
starwars |>
  filter(grepl('brown', hair_color) | eye_color == 'brown') |>
  select(name:eye_color, gender, homeworld, species) |>
  mutate(male_brunette = ifelse(gender == 'male' & hair_color == 'brown', TRUE, FALSE)) |>
  arrange(desc(height)) |>
  select(name, male_brunette, everything()) |>
  rename_all(str_replace, '_', ' ') |>
  rename_all(str_replace, 'color', 'colour') |>
  rename_all(str_to_title) |>
  print(n = 5)
```

```{r}
# by dropping the scoped variant (rename_all) and using the new rename_with
starwars |>
  filter(grepl('brown', hair_color) | eye_color == 'brown') |>
  select(name:eye_color, gender, homeworld, species) |>
  mutate(male_brunette = ifelse(gender == 'male' & hair_color == 'brown', TRUE, FALSE)) |>
  arrange(desc(height)) |>
  select(name, male_brunette, everything()) |>
  rename_with(~ str_replace(., '_', ' ')) |>
  rename_with(~ str_replace(., 'color', 'colour')) |>
  rename_with(str_to_title) |>
  print(n = 5)
```
