---
title: "Programming in R Homework Answers"
# Unlinked from the sidebar and excluded from site search so students do not
# stumble onto the answers. The page is still published, so the URL can be shared
# deliberately once the homework has been discussed.
search: false
author:
  - "Matthew Bracher-Smith"
  - "Gabriel Mateus Bernardo Harrington"
date: "2026-10-13"
format: 
  html:
    number-sections: true
    highlight: tango
editor: source
execute:
  echo: true
---

```{r}
#| output: false
#| warning: false
library(tidyverse)
library(conflicted)
conflicts_prefer(dplyr::filter)
```

This document contains all questions and answers for **Programming in R**. Please create a Quarto document containing all text, code and output used to answer the questions.

## Functions
1. Can you tell which of the arguments of myFunction are mandatory and which are optional?
```{r}
#| eval: false
myFunction <- function(x, y, verbose = FALSE){
  result <- x ** y
  if (verbose){
    print(result)
  }
  return(result)
}
```

_**x** and **y** are mandatory, but **verbose** is optional because a default value has already been assigned in the function definition._

2. What will be the output of the previous function if we type the following command myFunction(2, 3, TRUE)? Try to do it without running the code.

_There will be two outputs of 8. One is printed, because verbose is set to TRUE. The other is the return value of the function, which is automatically printed if we do not assign the output of a function to a variable._

3. Write `any_na()`, a function that takes two vectors of the same length and returns the number of positions that have an NA in, at least, one of the vectors.
```{r}
#| eval: false
# multiple options are available for ensuring the function takes arguments that are
# the same length. Extra points if they used a package like `assertthat`

# code below is what students might write to count all the NAs across both vectors
any_na <- function(x, y){
  stopifnot(length(x) == length(y))
  
  return(sum(c(is.na(x), is.na(y))))
}

# however, we want the number of positions, not number of total NAs
# again. There are many possible solutions. A simple one is to make a quick
# dataframe, as they have many functions available for NA operations
any_na <- function(x, y){
  # returns the number of positions that have an NA in at least one of the vectors
  stopifnot(length(x) == length(y))
  
  return(sum(!complete.cases(data.frame(x, y))))
}
```

_If any intermediate variables were created, they should have sensible names. A comment explaining what the function does is also welcome! Because we know nothing about the problem application for this question, something like x and y can be reluctantly accepted as names for input variables._

4. Write `complementary()`, a function that takes a DNA strand of variable length and returns its complementary strand.
```{r}
#| eval: false
# one option is using chartr, a function from base R built for character translation
complementary <- function(dna_seq){
  # returns complimentary strand for variable-length DNA strands
  return(chartr("ATGC","TACG", dna_seq))
}

# it wasn't specified what the strand would be in
# if we supplied the sequence as a vector, vectorised ifelse
# statement may be handy
complementary <- function(dna_seq){
  # returns complimentary strand for variable-length DNA strands
  return(dplyr::case_when(dna_seq == 'A' ~ 'T',
                          dna_seq == 'T' ~ 'A',
                          dna_seq == 'C' ~ 'G',
                          dna_seq == 'G' ~ 'C'))
}
```

_Whatever was used, variable and argument names should be appropriate. X or y are no longer acceptable, as we know something about the problem for this question._

5. Write a function that transcribes and translates a DNA strand into its corresponding protein sequence. You can assume that the length of the input DNA strand is a multiple of 3.

```{r}
#| eval: false
# students may have broken the problem into sub functions
# if this is the case, they should not be nested (they should
# not be defined within another function's body)
transcribe <- function(dna_seq){
  # converts from dna to rna
  rna_seq <- str_replace_all(dna_seq, 'T', 'U')
  
  return(rna_seq)
}

# even better than the option below would be putting the RNA codon table
# into a separate file which is loaded, so we can easily update it
translate <- function(rna_seq){
  # converts from RNA to protein
  # define an RNA codon table
  rna_codons = c("F", "L", "I", "V", "F", "L", "I", "V",
                "L", "L", "I", "V", "L", "L", "M", "V",
                "S", "P", "T", "A", "S", "P", "T", "A",
                "S", "P", "T", "A", "S", "P", "T", "A",
                "Y", "H", "N", "D", "Y", "H", "N", "D",
                "STOP", "Q", "K", "E", "STOP", "Q", "K", "E",
                "C", "R", "S", "G", "C", "R", "S", "G",
                "STOP", "R", "R", "G", "W", "R", "R", "G")
  names(rna_codons) <- c("UUU", "CUU", "AUU", "GUU", "UUC",
                         "CUC", "AUC", "GUC", "UUA", "CUA",
                         "AUA", "GUA", "UUG", "CUG", "AUG",
                         "GUG", "UCU", "CCU", "ACU", "GCU",
                         "UCC", "CCC", "ACC", "GCC", "UCA",
                         "CCA", "ACA", "GCA", "UCG", "CCG",
                         "ACG", "GCG", "UAU", "CAU", "AAU",
                         "GAU", "UAC", "CAC", "AAC", "GAC",
                         "UAA", "CAA", "AAA", "GAA", "UAG",
                         "CAG", "AAG", "GAG", "UGU", "CGU",
                         "AGU", "GGU", "UGC", "CGC", "AGC",
                         "GGC", "UGA", "CGA", "AGA", "GGA",
                         "UGG", "CGG", "AGG", "GGG" )
  
  # optionally check length is multiple of 3
  stopifnot((nchar(rna_seq) %% 3) == 0)
  
  # lookup one-letter amino acid code for each codon
  amino_acids <- ''
  for(i in seq(1, nchar(rna_seq), by=3)){
    codon <- substr(rna_seq, i, i+2)
    amino_acids <- paste(amino_acids, as.character(rna_codon[codon]), sep = '')
  }
  
  return(amino_acids)
}

dna_to_protein <- function(dna_seq) {
  # convert from dna to protein
  rna_seq <- transcribe(dna_seq)
  protein_seq <- translate(rna_seq)
  return(protein_seq)
}
```

_As before, variable names and function names should be appropriate. Comments are probably needed to make parts clear. It makes sense for separate operations to be broken into separate functions like above. This format is much-preferred. If anyone did the whole thing in a single function, some effort should be made to clearly separate-out the different sections._
## Writing a cleaning function

```{r}
#| message: false
baseline <- read_csv(
  here::here("Simulated_data", "generated_uk", "bioinformatics_msc_stats_baseline.csv"),
  show_col_types = FALSE
)

baseline |> count(obesity)
```

6. Write `clean_yes_no()`, returning a logical vector.

```{r}
clean_yes_no <- function(x) {
  v <- str_to_lower(str_trim(x))
  unrecognised <- setdiff(unique(v[!is.na(v)]), c("yes", "y", "no", "n"))
  if (length(unrecognised) > 0) {
    stop("Unrecognised value(s): ", paste(unrecognised, collapse = ", "), call. = FALSE)
  }
  case_when(v %in% c("yes", "y") ~ TRUE, v %in% c("no", "n") ~ FALSE, .default = NA)
}
```

Note what the function does *not* do: it does not guess. `str_to_lower()` and
`str_trim()` handle the variation we expect — case and stray whitespace —
and anything else is an error rather than a silent `NA`.

### 6a. Failing loudly

```{r}
#| error: true
clean_yes_no(c("Yes", "No", "maybe"))
```

Returning `NA` for unrecognised input would mean a new coding error — a `"Y "`,
a `"1"`, a `"Unknown"` — silently becomes missing data. The row count stays the
same, nothing warns you, and your prevalence is quietly wrong. Erring forces
somebody to look at it and decide.

### 6b. Prevalence

```{r}
obese <- clean_yes_no(baseline$obesity)

c(n = length(obese), n_obese = sum(obese), prevalence = round(mean(obese), 3))
```

33.8%, which matches the figure quoted in the dataset README — a useful check
that the cleaning did what we intended rather than something plausible-looking.

### 6c. Reuse

```{r}
t2d <- clean_yes_no(baseline$type2_diabetes)
round(mean(t2d), 3)
```

It works unchanged, and gives 8.2% — again matching the README. That is the
argument for a function over a one-off fix: the same rule now applies to every
yes/no column in the dataset, it is testable, and if the rule turns out to be
wrong you change it in one place.

7. `check_range()`

```{r}
check_range <- function(x, min, max) {
  which(!is.na(x) & (x < min | x > max))
}
```

### 7a. `imd_quintile`

```{r}
bad_imd <- check_range(baseline$imd_quintile, 1, 5)

baseline[bad_imd, c("participant_id", "imd_quintile")]
```

One participant is coded `9`, which is not a valid quintile. Most likely a
missing-value code that was never translated on import — `9` and `99` are
conventional "not known" codes in older clinical datasets.

### 7b. `systolic_bp_mmHg`

```{r}
bad_bp <- check_range(baseline$systolic_bp_mmHg, 60, 250)

baseline[bad_bp, c("participant_id", "age_years", "systolic_bp_mmHg")]
```

`1330` and `1620` are `133.0` and `162.0` with a misplaced decimal point — both
entirely plausible blood pressures, recorded ten times too large. Note that the
range you choose is a judgement: too wide and you miss them, too narrow and you
flag real outliers. State the range you used and why.
